Modeling tabular icebergs coupled to an ocean model

### 2017-03-29 JAMES_iceberg_only.tex

\begin{keypoints}
\item A novel modeling framework is developed to explicitly model large tabular icebergs submerged in the ocean.
\item Tabular icebergs are constructed out of Lagrangian elements that drift in the ocean, and are held together by numerical bonds.
%\item The model capabilities are demonstrated by modeling a tabular iceberg drifting away an idealized ice shelf.
\item Breaking the numerical bonds allows us to model iceberg breakup and calving.
\end{keypoints}

@@

\begin{abstract}
Large tabular icebergs calved from Antarctic ice shelves have long lifetimes (due to their large size), during which they drift across large distances, altering regional ocean circulation, bottom-water formation, sea-ice formation and biological primary productivity in the icebergs' vicinity.
However, despite their importance, the current generation of ocean circulation models is unable to represent large tabular icebergs.
In this study we develop a novel framework to model large tabular icebergs submerged in the ocean. In this framework, tabular icebergs are composed of Lagrangian elements that drift in the ocean, and are held together and interact with each other via bonds.
%We describe the Lagrangian bonded iceberg model and discuss a number of the technical elements involved in its development. 
%We demonstrate the capabilities of the model by modeling a tabular iceberg drifting away an idealized ice shelf.} 
A break of these bonds allows the model to emulate calving events (i.e. detachment of a tabular iceberg from an ice shelf), and tabular icebergs breaking up into smaller pieces. Idealized simulations of a calving tabular iceberg, its drift, and of a tabular iceberg breaking into two pieces, demonstrate capabilities of the developed framework.
\end{abstract}

@@

After calving, icebergs slowly drift away from their origins, often becoming stuck in sea ice, or grounding on bathymetric highs along the Antarctic coast \citep{Lichey2001, Dowdeswell2007}. Large tabular icebergs extend deep into the water column, and have the potential to disrupt ocean circulation patterns for months or even years after calving \citep{Robinson2012, Stern2015}.
The freshwater flux from iceberg melt impacts ocean hydrography around the iceberg, influencing sea-ice production and bottom-water formation \citep{Arrigo2002, Robinson2012, Nicholls2009, Fogwill2016}. Because of their large size, the tabular icebergs have long lifetimes during which they drift over long distances injecting meltwater along the way and impacting the Southern Ocean state (e.g. hydrography, sea ice conditions, etc.) far away from their calving origins \citep{Stern2016}. Meltwater injection (and the accompanying upwelling) from icebergs can also influence biological productivity by bringing nutrients to the surface ocean or changing sea ice conditions \citep{Arrigo2002, Vernet2012, Biddle2015}. The increased productivity associated with free-floating tabular icebergs has been linked with local increases in ocean carbon uptake, potentially large enough to be a significant fraction of the Southern Ocean carbon sequestration \citep{Smith2007}.

@@

In recent years, there has been an increased interest in iceberg drift and decay. This surge of interest has been driven by (i) the need to understand polar freshwater cycles in order to create realistic climate forecasts and sea level projections \citep{Silva2006, Shepherd2007, Rignot2013}; and (ii) the increased navigation and exploration activities in high-latitude iceberg-filled waters in the Arctic \citep{Pizzolato2014, Unger2014, Henderson2016}. The increased interest in icebergs has led to the development of numerical models of iceberg drift and decay \citep{Mountain1980, Bigg1997, Gladstone2001, Kubat2005}, some of which have been included in global General Circulation Models \citep{Martin2010, Marsh2015}. These iceberg drift models treat icebergs as Lagrangian point particles, which are advected by the flow, and melt according to certain parameterizations for icebergs melt. Since icebergs are treated as point particles, iceberg drift models are mostly suitable for modeling icebergs smaller than an ocean grid cell. Consequently, these models have mostly been used to represent icebergs smaller than 3.5~km on a global scale \citep{Jongma2009, Martin2010, Marsh2015}.

Point-particle iceberg drift models are less suitable for modeling larger tabular icebergs, where the size and structure of the iceberg may be an important feature in determining their drift and decay \citep{Stern2016}. They also are not suitable for studying the local effects that icebergs have on the surrounding ocean, or the small scale processes that influence iceberg melt and decay \citep{Wagner2014, Stern2015}. For these reasons, tabular icebergs are currently not represented in the iceberg drift models used as components of climate models, despite accounting for the vast majority of the total Southern Hemisphere iceberg mass \citep{Tournadre2016}. Point-particle iceberg models also do not have any representation of iceberg breakup and calving, which is know to be an important iceberg decay mechanism and influences iceberg trajectories.
%Similarly, operational iceberg models are often inadequate for projecting iceberg drift at fine scales \citep{Turnbull2015}.

The goal of this study is to develop a new framework to model all kinds of icebergs, where tabular icebergs are explicitly resolved in the ocean. Our new representation of icebergs aims to include the following key properties: (i) icebergs should be able to travel large distances within the ocean, (ii)  icebergs should melt and decay as they drift in the ocean, (iii) icebergs should behave as if they has finite extent (so that we can use the model to study local effects that icebergs have on the surrounding ocean), and (iv) tabular icebergs should be able to break away from ice shelves or break into smaller pieces. Properties (i) and (ii) are common to point-particle icebergs models, while properties (iii) and (iv) are new to the framework developed in this study. A further requirement of the new framework is that the model should run sufficiently quickly to be used in general circulation models used for climate.

In order to allow icebergs to travel large distances, it is natural to model the icebergs in a Lagrangian framework (as in the point particle iceberg drift models described above). In our model we therefore represent icebergs using Lagrangian elements. However in our model icebergs are no longer treated as point particles that interact with the ocean at a single location, but rather icebergs are given physical structure, so that they interact with the ocean across multiple ocean grid cells, depress the ocean surface over a wide area, and can interact with other icebergs  (see schematic in Figure \ref{fig:Hex_schematic}). This is done by assigning a finite surface area and shape to the Lagrangian elements, which allows the elements to behave as if they has a finite extent. 
%Element interact with the ocean by distributing their weight, surface area and melt fluxes over multiple ocean grid cells in a way which is consistent with the shape of the ice element.
The finite extent of an element is transmitted by the ocean by distributing the element's weight, surface area and melt fluxes over multiple ocean grid cells in a way which is consistent with the shape of the ice element. The finite extent of an element is felt by other elements through repulsive forces which are applied when the boundaries of the elements overlap.
%Elements interact with other elements when the boundaries of the elements overlap (see schematic in Figure \ref{fig:Hex_schematic}).

Large tabular icebergs can then be represented by `bonding' together multiple ice elements using numerical bonds (see schematic in Figure \ref{fig:Hex_schematic}).
The numerical bonds hold the ice elements together and allow a collection of elements to move as a unit. 
This allows us to study how tabular icebergs drift in the ocean when forced by ocean currents and wind.
%Figure \ref{fig:Hex_schematic}) shows a schematic of the difference between point-particle icebergs, finite extent icebergs and tabular icebergs constructed out of multiple elements. 
%Figure \ref{fig:Hex_schematic} shows a schematic of point-particle elements, elements with finite extent and bonded elements, and shows schematically how tabular icebergs can be constructed out of multiple elements.
An advantage of representing tabular icebergs using numerical bonds is that by manually breaking the bonds, we can simulate iceberg calving, allowing us to study the ocean response to tabular icebergs detaching from ice shelves (e.g.: Figure \ref{fig:SST_Collapse}), or the response to an iceberg fracturing into multiple smaller pieces.

@@

The manuscript is organized as follows. 
Section 2 gives a description of many of the key aspects of the model developed in this study. Since this model is a new approach to modeling icebergs, we present many of the technical aspects of the model.  In Sections 3 and 4, we demonstrate the capabilities of the model by simulating a tabular iceberg detaching from an idealized ice shelf. In a further simulation we break some numerical bonds within the tabular iceberg to demonstrate an iceberg splitting in two.

@@

The Lagrangian Bonded Iceberg Model (LBIM) is a Lagrangian particle-based model in that the objects of the model are Lagrangian elements. Each element represents a column of ice that is floating in the ocean, and has a position, velocity, mass, and a set of dimensions, which can evolve in time. The motion of each element is determined by a momentum equation which is solved in the (Lagrangian) reference frame of the element. The elements experience oceanic and atmospheric forces, which are either prescribed, or computed by coupling the LBIM to an ocean/atmosphere model. The LBIM elements also interact with one another and can be bonded together to form larger structures. The angular momentum of the elements is not modeled explicitly; instead rotational motion of larger structures emerge as a consequence of bond orientation and collective motion.

@@

The LBIM is developed on the code base of an existing iceberg drift model \citep{Martin2010, Stern2016}. When run with the correct set of runtime flags, the model runs as a traditional iceberg drift model.
%
%In this section, we discuss the equations of motion governing the element movement and the thermodynamic equations governing the decay of the elements. The also discuss the interactive forces

\subsection{Equations of motion}
The elements drift in the ocean, in response to atmosphere, ocean and sea-ice drag forces, as well as the Coriolis force and a force due to the sea surface slope.  When these ice elements move alone (without interactions with other elements), they can be thought of as representing individual (or clusters of) small icebergs, and follow the same equations described in the iceberg drift model of \citet{Martin2010} (based on the equations outlined in \citet{Bigg1997} and \citet{Gladstone2001}).

In addition to the external forces, the elements in the LBIM experience interactive forces due to the presence of other elements. Two types of interactive forces are included between elements. The first force is a repulsive force which is applied to elements to prevent them from overlapping the boundaries of the neighboring elements. The second interactive force is a force due to numerical `bonds', and is only applied if two elements are labelled as `bonded'. When two elements are bonded, each element feels an attractive force that prevents the elements from moving too far apart from one another. 
%The interactive forces between two bonded elements are defined such that in the absence of other forces the elements come to rest adjacent to one another, with no overlap of the iceberg areas. \textcolor{red}{???The last sentence is unclear???}

The momentum equation for each element is given by 
\begin{equation}
\label{Iceberg_mom_eq}
M \frac{D \vec{u}}{Dt}=\vec{F}_{A} + \vec{F}_{W}+ \vec{F}_{R}+\vec{F}_{C} +\vec{F}_{SS}+\vec{F}_{SI}+\vec{F}_{IA} ,
\end{equation}
where $\frac{D}{Dt}$ is the total (Lagrangian) derivative, $M$ is the mass of the element, $\vec{u}$ is the velocity of the element, and the terms on the right hand side give the forces on the element due to air drag $(\vec{F}_{A})$, water drag $(\vec{F}_{W})$, sea ice drag $(\vec{F}_{SI})$, Coriolis force $(\vec{F}_{C})$, wave radiation force $(\vec{F}_{R})$, sea surface slope $(\vec{F}_{SS})$, and interactions with other elements ($\vec{F}_{IA}$). The environmental forces are the same as those presented in \citet{Martin2010}, and are provided for completeness in Appendix A.  The details of the interactive forces are provided in below.

\subsection{Interactive Forces} 
%As described above, our model treats icebergs as rigid icebergs with finite extent, which move through the ocean.
The interactive forces in the LBIM model are used to (i) prevent the ice elements from overlapping and (ii) to connect multiple ice elements together so that the collection of elements moves as a rigid body.
Modeling the collisions and movements of rigid objects precisely, requires very small time steps and precise collision detection algorithms, which are very computationally expensive. Models using these methods are typically only run for a few days or even a few seconds, and are used to study rapid processes like crack formation or ridging \citep{Hopkins2004, Bassis2013, Rabatel2015}). The tabular iceberg framework presented in this study is developed in order to be used in general circulation models used for climate, and therefore must be run for many years. In order to gain the required computational efficiency, we relax the requirement that icebergs must be perfectly rigid and that ice elements can not overlap. Instead, we model the interactive forces between ice elements using damped elastic forces, which can be calculated more rapidly. %A consequence of this is that the tabular icebergs can bend slightly, and ice elements can overlap slightly. However, these effects are sufficiently small that it

@@

To describe the forces between two elements, we begin by introducing some notation.  Let $\vec{x_{i}}$, $\vec{x_{j}}$ be the positions of elements $i$ and $j$. The distance between elements $i$ and $j$ is
\begin{equation}
\label{distance}
d_{ij}= | \vec{x_{i}} - \vec{x_{j}}|.
\end{equation}
When calculating the interactive forces between elements, the elements are assumed to be circular. We define the interaction radius of an element by 
\begin{equation}
\label{Diameter}
R_{i}=\sqrt{\frac{A_{i}}{\pi}},
\end{equation}
where $A_{i}$ is the planar surface area of element $i$. Using this, we define the critical-interactive-length scale, 
\begin{equation}
L_{ij}=R_{i} + R_{j},
\label{eq:Interaction_crit}
\end{equation}
which governs interactions between elements $i$ and $j$. 
Repulsive forces are only applied when $d_{i,j} \textless L_{i,j}$, while for $d_{i,j} \textgreater L_{i,j}$ attractive bonded forces are applied when a bond exists between element $i$ and $j$ (see diagram in Figure \ref{fig:Force_diagram}). 
The interactive forces are designed such that (in the absence of other external forces) bonded particles will settle in an equilibrium position where elements are separated by $L_{i,j}$.

@@

To aid in notation, we define a bond matrix $B_{ij}$ such that $B_{ij}=1$ if elements $i$ and $j$ are bonded together and  $B_{ij}=0$ otherwise. Using this notation, the interactive force $(\vec{F}_{IA})_{ij}$ on an element $i$ by an element $j$ is given by
\begin{equation}
\label{F_interactione}
(\vec{F}_{IA})_{ij}= \left\{%
  \begin{array}{ll}
   (\vec{F}_{e})_{ij} + (\vec{F}_{d})_{ij}
    & \mbox{if $d_{ij} \leq L_{ij} $ }\\
 %     & \mbox{if $|x_{i} -x_{j}| \leq \frac{1}{2}(D_{i}+D_{j}) $ }\\
    (\vec{F}_{e})_{ij} + (\vec{F}_{d})_{ij}
 %     & \mbox{if $|x_{i} -x_{j}| > \frac{1}{2}(D_{i}+D_{j}) \textrm{ and } B_{ij}=1 $ }\\
       & \mbox{if $d_{ij} >  L_{ij}  \textrm{ and } B_{ij}=1 $ }\\
      0
 %     & \mbox{if $|x_{i} -x_{j}| > \frac{1}{2}(D_{i}+D_{j}) \textrm{ and } B_{ij}=0 $ }.\\
      & \mbox{if $d_{ij} > L_{ij} \textrm{ and } B_{ij}=0 $ }.\\
  \end{array}
\right.
\end{equation}
$(\vec{F_{e}})_{ij}$ and $(\vec{F_{d}})_{ij}$ are the elastic and frictional damping components of the interactive force between elements $i$ and $j$.
The elastic force $(\vec{F_{e}})_{ij}$ between elements is given by
\begin{equation}
\label{F_elastic}
(\vec{F_{e}})_{ij}= -\kappa_{e}  \bigg( d_{i,j}-L_{i,j} \bigg) T_{i,j} \vec{r}_{ij},
\end{equation}
where $\vec{r}_{ij}= \frac{(\vec{x_{i}} -\vec{x_{j}})}{|\vec{x_{i}} -\vec{x_{j}}|}$ is the directional unit vector between the position of element $i$ and $j$, $\kappa_{e}$ is the spring constant, and $T_{i,j}$ is the minimum of the thickness of elements $i$ and $j$. The interactive forces obey Newton's 3rd Law (i.e.: $(\vec{F}_{IA})_{ij}=-(\vec{F}_{IA})_{ji}$).
%This is in contrast to some other DEM models in the literature (e.g. \citet{Li2014}). 
The minimum thickness, $T_{i,j}$, is preferred to the average thickness, since this means that for two bonded elements a fixed distance apart, the acceleration due to elastic forces is bounded, even when the thickness of one of the elements approaches zero.

@@

The frictional damping force acts to oppose the relative velocity of the two particles.
%, both in the direction parallel to $\vec{r}_{ij}$ and in the direction perpendicular to  $\vec{r}_{ij}$.
If $\vec{r}_{ij}^{\perp}$ is the direction vector perpendicular to $\vec{r}_{ij}$, and $P_{\vec{r}_{ij}}$ and $P_{\vec{r}_{ij}^{\perp}}$ are the projection matrices that project onto $\vec{r}_{ij}$ and $\vec{r}_{ij}^{\perp}$ respectively, then the frictional damping force is given by
\begin{equation}
\label{F_damping} 
%(F_{d})_{ij}=   -C_{r}|(u_{i} -u_{j}) \cdot \vec{r}   | \vec{r}  - C_{t}|(u_{i} -u_{j})\cdot  \vec{r}^{\perp} | \vec{r}^{\perp})
(\vec{F_{d}})_{ij}=   \Bigg (-c_{r_{||}} P_{\vec{r}_{ij}}  - c_{r_{\perp}} P_{\vec{r}_{ij}^{\perp}} \Bigg ) \cdot (\vec{u_{i}} -\vec{u_{j}} ) 
\end{equation}
Here $c_{r_{||}}$ and $c_{r_{\perp}}$ are the drag coefficients for the damping parallel and perpendicular to $r_{ij}$, respectively. We set $c_{r_{||}}=2\sqrt{\kappa_{e} }$, so that the elastic force parallel to $\vec{r}_{ij}$ is critically damped. The damping coefficient in the the perpendicular direction is set to $c_{r_{\perp}}=\frac{1}{4}c_{r_{||}}$. 
%The frictional damping force perpendicular to $\vec{r}_{ij}$ helps to reduce the motion of particles when many particles are packed tightly together (as in the uncoupled experiments below), and also acts to reduce the vibrations within larger bonded structures. 
The damping forces are implemented using an implicit time stepping scheme, to avoid stability issues for very small elements (details found in Appendix B).

>>>>

\begin{keypoints}
\item A novel modeling framework is developed to explicitly model large tabular icebergs submerged in the ocean.
\item Tabular icebergs are represented using Lagrangian elements that drift in the ocean, and are held together by numerical bonds.
%\item The model capabilities are demonstrated by modeling a tabular iceberg drifting away an idealized ice shelf.
\item Breaking the numerical bonds allows us to model iceberg breakup and calving.
\end{keypoints}

@@

\begin{abstract}
Large tabular icebergs calved from Antarctic ice shelves have long lifetimes (due to their large size), during which they drift across large distances, altering ambient ocean circulation, bottom-water formation, sea-ice formation and biological primary productivity in the icebergs' vicinity.
However, despite their importance, the current generation of ocean circulation models do not represent large tabular icebergs.
In this study we develop a novel framework to model large tabular icebergs submerged in the ocean. In this framework, tabular icebergs are represented by Lagrangian elements that drift in the ocean, and are held together and interact with each other via bonds.
%We describe the Lagrangian bonded iceberg model and discuss a number of the technical elements involved in its development. 
%We demonstrate the capabilities of the model by modeling a tabular iceberg drifting away an idealized ice shelf.} 
A break of these bonds allows the model to emulate calving events (i.e. detachment of a tabular iceberg from an ice shelf) and tabular icebergs breaking up into smaller pieces. Idealized simulations of a calving tabular iceberg, its drift, and breaking into two pieces, demonstrate capabilities of the developed framework.
\end{abstract}

@@

After calving, icebergs drift away from their origins, often becoming stuck in sea ice, or grounding on bathymetric highs along the Antarctic coast \citep{Lichey2001, Dowdeswell2007}. Large tabular icebergs extend deep into the water column, and have the potential to disrupt ocean circulation patterns for months or even years after calving \citep{Robinson2012, Stern2015}.
The freshwater flux from iceberg melt impacts ocean hydrography around the iceberg, influencing sea-ice production and bottom-water formation \citep{Arrigo2002, Robinson2012, Nicholls2009, Fogwill2016}. Because of their large size, the tabular icebergs have long lifetimes during which they drift over long distances injecting meltwater along the way and impacting the Southern Ocean state (e.g. hydrography, sea ice conditions, etc.) far away from their calving origins \citep{Stern2016, Rackow2017}. Meltwater injection (and the accompanying upwelling) from icebergs can also influence biological productivity by bringing nutrients to the surface ocean or changing sea ice conditions \citep{Arrigo2002, Vernet2012, Biddle2015}. The increased productivity associated with free-floating tabular icebergs has been linked with local increases in ocean carbon uptake, potentially large enough to be a significant fraction of the Southern Ocean carbon sequestration \citep{Smith2007}.

@@

In recent years, there has been an increased interest in iceberg drift and decay. This surge of interest has been driven by (i) the need to understand polar freshwater cycles in order to create realistic climate change and sea level projections \citep{Silva2006, Shepherd2007, Rignot2013}; and (ii) the increased navigation and exploration activities in high-latitude iceberg-filled waters in the Arctic \citep{Pizzolato2014, Unger2014, Henderson2016}. The increased interest in icebergs has led to the development of numerical models of iceberg drift and decay \citep{Mountain1980, Bigg1997, Gladstone2001, Kubat2005}, some of which have been included in global General Circulation Models \citep{Martin2010, Marsh2015}. These iceberg drift models treat icebergs as Lagrangian point particles, which are advected by the flow, and melt according to parameterizations for icebergs melt. Since icebergs are treated as point particles, iceberg drift models are mostly suitable for modeling icebergs smaller than an ocean grid cell. Consequently, these models have mostly been used to represent icebergs smaller than 3.5~km on a global scale \citep{Jongma2009, Martin2010, Marsh2015}.

Point-particle iceberg drift models are less suitable for modeling larger tabular icebergs, where the size and structure of the iceberg may be an important feature in determining their drift and decay \citep{Stern2016}. They also are not suitable for studying the local effects that icebergs have on the surrounding ocean, or the small scale processes that influence iceberg melt and decay \citep{Wagner2014, Stern2015}. For these reasons, tabular icebergs are currently not represented in the iceberg drift models used as components of climate models, despite accounting for the vast majority of the total Southern Hemisphere iceberg mass \citep{Tournadre2016}. Point-particle iceberg models also do not have any representation of iceberg breakup and calving, which is known to be an important iceberg decay mechanism that influences iceberg trajectories.
%Similarly, operational iceberg models are often inadequate for projecting iceberg drift at fine scales \citep{Turnbull2015}.

The goal of this study is to develop a new framework to model all kinds of icebergs, where tabular icebergs are explicitly resolved in the ocean. Our new representation of icebergs aims to include the following key properties: (i) icebergs should be able to travel large distances within the ocean, (ii)  icebergs should melt and decay as they drift in the ocean, (iii) icebergs should behave as if they have finite extent (in order to study local effects that icebergs have on the surrounding ocean), and (iv) tabular icebergs should be able to break away from ice shelves or break into smaller pieces. Properties (i) and (ii) are common to point-particle icebergs models, while properties (iii) and (iv) are new to the framework developed in this study. A further requirement of the new framework is that the model should run sufficiently quickly to be used in general circulation models used for climate.

In order to allow icebergs to travel large distances, we model the icebergs in a Lagrangian framework (as in the point particle iceberg drift models described above). However in our model icebergs are no longer treated as point particles that interact with the ocean at a single location. Instead icebergs are given physical structure, so that they interact with the ocean across multiple ocean grid cells, depress the ocean surface over a wide area, and can interact with other icebergs (Figure \ref{fig:Hex_schematic}). This is done by assigning a finite surface area and shape to the Lagrangian elements, which allows the elements to behave as if they have a finite extent. 
%Element interact with the ocean by distributing their weight, surface area and melt fluxes over multiple ocean grid cells in a way which is consistent with the shape of the ice element.
The finite extent of an element is transmitted by the ocean by distributing the element's weight, surface area and melt fluxes over multiple ocean grid cells in a way which is consistent with the shape of the ice element. 
Finite-extent elements interact with each other via repulsive forces which are applied when the boundaries of the elements overlap.
%Elements interact with other elements when the boundaries of the elements overlap (see schematic in Figure \ref{fig:Hex_schematic}).

Large tabular icebergs can then be represented by `bonding' together multiple ice elements into larger structures using numerical bonds (Figure \ref{fig:Hex_schematic}).
The numerical bonds hold the ice elements together and allow a collection of elements to move as a unit. 
This allows tabular icebergs to drift in the ocean when forced by ocean currents and wind.
%Figure \ref{fig:Hex_schematic}) shows a schematic of the difference between point-particle icebergs, finite extent icebergs and tabular icebergs constructed out of multiple elements. 
%Figure \ref{fig:Hex_schematic} shows a schematic of point-particle elements, elements with finite extent and bonded elements, and shows schematically how tabular icebergs can be constructed out of multiple elements.
An advantage of representing tabular icebergs using numerical bonds is that by breaking the bonds, we can simulate iceberg calving (e.g.: Figure \ref{fig:SST_Collapse}), or the response to an iceberg fracturing into multiple smaller pieces.

@@

The manuscript is organized as follows. 
Section 2 gives a description of the key aspects of the model developed in this study. Since this model is a new approach to modeling icebergs, we present technical aspects of the model.  In Sections 3 and 4, we demonstrate the capabilities of the model by simulating a tabular iceberg detaching from an idealized ice shelf. In a further simulation we break some numerical bonds within the tabular iceberg to demonstrate an iceberg splitting in two.

@@

The Lagrangian Bonded Iceberg Model (LBIM) is a Lagrangian particle-based model in that the objects of the model are Lagrangian elements. Each element represents a column of ice that is floating in the ocean, and has a position, velocity, mass, and a set of dimensions, which can evolve in time. The motion of each element is determined by a momentum equation which is solved in the (Lagrangian) reference frame of the element. The elements experience oceanic and atmospheric forces, which are either prescribed, or computed by coupling the LBIM to an ocean/atmosphere model. The LBIM elements also interact with one another via attractive and repulsive interactive forces, and can be bonded together to form larger structures. The angular momentum of the elements is not modeled explicitly; instead rotational motion of larger structures emerge as a consequence of bond orientation and collective motion.

@@

The LBIM is developed on the code base of an existing iceberg drift model \citep{Martin2010, Stern2016}. When run with the correct set of runtime flags, the model runs as a traditional point-particle iceberg drift model.
%
%In this section, we discuss the equations of motion governing the element movement and the thermodynamic equations governing the decay of the elements. The also discuss the interactive forces

\subsection{Equations of motion}
The elements drift in the ocean in response to atmosphere, ocean and sea-ice drag forces, as well as the Coriolis force, a wave radiation force, a force due to the sea surface slope and interactive forces with other elements. The momentum equation for each element is given by 
\begin{equation}
\label{Iceberg_mom_eq}
M \frac{D \vec{u}}{Dt}=\vec{F}_{A} + \vec{F}_{W}+ \vec{F}_{R}+\vec{F}_{C} +\vec{F}_{SS}+\vec{F}_{SI}+\vec{F}_{IA} ,
\end{equation}
where $\frac{D}{Dt}$ is the total (Lagrangian) derivative, $M$ is the mass of the element, $\vec{u}$ is the velocity of the element, and the terms on the right hand side give the forces on the element due to air drag $(\vec{F}_{A})$, water drag $(\vec{F}_{W})$, sea ice drag $(\vec{F}_{SI})$, Coriolis force $(\vec{F}_{C})$, wave radiation force $(\vec{F}_{R})$, sea surface slope $(\vec{F}_{SS})$, and interactions with other elements ($\vec{F}_{IA}$).

% 
When ice elements move alone (without interactions with other elements), they can be thought of as representing individual (or clusters of) small icebergs, and follow the same equations described in the iceberg drift model of \citet{Martin2010} (based on the equations outlined in \citet{Bigg1997} and \citet{Gladstone2001}).
These environmental forces are provided for completeness in Appendix A.

In addition to the external forces, the elements in the LBIM experience interactive forces due to the presence of other elements. Two types of interactive forces are included between elements. The first force is a repulsive force which is applied to elements to prevent them from overlapping the boundaries of the neighboring elements. The second interactive force is a force due to numerical `bonds', and is only applied if two elements are labelled as `bonded'. When two elements are bonded, each element feels an attractive force that prevents the elements from moving too far apart from one another.  The details of the interactive forces are provided in below. 
%The interactive forces between two bonded elements are defined such that in the absence of other forces the elements come to rest adjacent to one another, with no overlap of the iceberg areas. \textcolor{red}{???The last sentence is unclear???}

\subsection{Interactive Forces} 
%As described above, our model treats icebergs as rigid icebergs with finite extent, which move through the ocean.
The interactive forces in the model are used to (i) prevent the ice elements from overlapping and (ii) to connect multiple ice elements together so that the collection of elements moves as a rigid body.
Modeling the collisions and movements of rigid objects precisely, requires very small time steps and precise collision detection algorithms, which are very computationally expensive. Models using these methods are typically only run for a few days or even a few seconds, and are used to study rapid processes like crack formation or ridging \citep{Hopkins2004, Bassis2013, Rabatel2015}). The tabular iceberg framework presented in this study is developed in order to be used in general circulation models used for multi-year simulations. In order to gain the required computational efficiency, we relax the requirement that icebergs must be perfectly rigid and that ice elements can not overlap. Instead, we model the interactive forces between ice elements using damped elastic forces, which can be calculated more efficiently. %A consequence of this is that the tabular icebergs can bend slightly, and ice elements can overlap slightly. However, these effects are sufficiently small that it

@@

To describe the forces between two elements, we begin by introducing some notation.  Let $\vec{x_{i}}$, $\vec{x_{j}}$ be the positions of elements $i$ and $j$. The distance between elements $i$ and $j$ is
\begin{equation}
\label{distance}
d_{ij}= | \vec{x_{i}} - \vec{x_{j}}|.
\end{equation}
In calculations of the interactive forces between elements, the elements are assumed to be circular. We define the interaction radius of an element by 
\begin{equation}
\label{Diameter}
R_{i}=\sqrt{\frac{A_{i}}{\pi}},
\end{equation}
where $A_{i}$ is the planar surface area of element $i$. Using this, we define the critical-interactive-length scale, 
\begin{equation}
L_{ij}=R_{i} + R_{j},
\label{eq:Interaction_crit}
\end{equation}
which governs interactions between elements $i$ and $j$. 
Repulsive forces are only applied when $d_{i,j} \textless L_{i,j}$, while for $d_{i,j} \textgreater L_{i,j}$ attractive bonded forces are applied when a bond exists between element $i$ and $j$ (see diagram in Figure \ref{fig:Force_diagram}). 
The interactive forces are designed such that (in the absence of other external forces) bonded particles will settle in an equilibrium position where elements are separated by $L_{i,j}$.

@@

To aid in notation, we define a bond matrix $B_{ij}$ such that $B_{ij}=1$ if elements $i$ and $j$ are bonded together and  $B_{ij}=0$ otherwise. Using this notation, the interactive force $(\vec{F}_{IA})_{ij}$ on an element $i$ by an element $j$ is given by
\begin{equation}
\label{F_interactione}
(\vec{F}_{IA})_{ij}= \left\{%
  \begin{array}{ll}
   (\vec{F}_{e})_{ij} + (\vec{F}_{d})_{ij}
    & \mbox{if $d_{ij} \leq L_{ij} $ }\\
 %     & \mbox{if $|x_{i} -x_{j}| \leq \frac{1}{2}(D_{i}+D_{j}) $ }\\
    (\vec{F}_{e})_{ij} + (\vec{F}_{d})_{ij}
 %     & \mbox{if $|x_{i} -x_{j}| > \frac{1}{2}(D_{i}+D_{j}) \textrm{ and } B_{ij}=1 $ }\\
       & \mbox{if $d_{ij} >  L_{ij}  \textrm{ and } B_{ij}=1 $ }\\
      0
 %     & \mbox{if $|x_{i} -x_{j}| > \frac{1}{2}(D_{i}+D_{j}) \textrm{ and } B_{ij}=0 $ }.\\
      & \mbox{if $d_{ij} > L_{ij} \textrm{ and } B_{ij}=0 $ }.\\
  \end{array}
\right.
\end{equation}
$(\vec{F_{e}})_{ij}$ and $(\vec{F_{d}})_{ij}$ are the elastic and frictional damping components of the interactive force between elements $i$ and $j$.
The elastic force $(\vec{F_{e}})_{ij}$ between elements is given by
\begin{equation}
\label{F_elastic}
(\vec{F_{e}})_{ij}= -\kappa_{e}  \bigg( d_{i,j}-L_{i,j} \bigg) M_{i,j} \vec{r}_{ij},
\end{equation}
where $\vec{r}_{ij}= \frac{(\vec{x_{i}} -\vec{x_{j}})}{|\vec{x_{i}} -\vec{x_{j}}|}$ is the directional unit vector between the position of element $i$ and $j$, $\kappa_{e}$ is the spring constant, and $M_{i,j}$ is the minimum of the masses of elements $i$ and $j$. The interactive forces obey Newton's 3rd Law (i.e.: $(\vec{F}_{IA})_{ij}=-(\vec{F}_{IA})_{ji}$).
%This is in contrast to some other DEM models in the literature (e.g. \citet{Li2014}). 
The minimum mass, $M_{i,j}$, is preferred to the average thickness, since this means that for two bonded elements a fixed distance apart, the acceleration due to elastic forces is bounded, even when the thickness of one of the elements approaches zero.

@@

The frictional damping force acts to dampen the relative motion of the two particles.
%, both in the direction parallel to $\vec{r}_{ij}$ and in the direction perpendicular to  $\vec{r}_{ij}$.
If $\vec{r}_{ij}^{\perp}$ is the direction vector perpendicular to $\vec{r}_{ij}$, and $P_{\vec{r}_{ij}}$ and $P_{\vec{r}_{ij}^{\perp}}$ are the projection matrices that project onto $\vec{r}_{ij}$ and $\vec{r}_{ij}^{\perp}$ respectively, then the frictional damping force is given by
\begin{equation}
\label{F_damping} 
%(F_{d})_{ij}=   -C_{r}|(u_{i} -u_{j}) \cdot \vec{r}   | \vec{r}  - C_{t}|(u_{i} -u_{j})\cdot  \vec{r}^{\perp} | \vec{r}^{\perp})
(\vec{F_{d}})_{ij}=   \Bigg (-c_{r_{||}} P_{\vec{r}_{ij}}  - c_{r_{\perp}} P_{\vec{r}_{ij}^{\perp}} \Bigg ) \cdot (\vec{u_{i}} -\vec{u_{j}} ) 
\end{equation}
Here $c_{r_{||}}$ and $c_{r_{\perp}}$ are the drag coefficients for the damping motion parallel and perpendicular to $r_{ij}$, respectively. We set $c_{r_{||}}=2\sqrt{\kappa_{e} }$, so that the elastic force parallel to $\vec{r}_{ij}$ is critically damped. The perpendicular drag coefficient is set to $c_{r_{\perp}}=\frac{1}{4}c_{r_{||}}$. The perpendicular damping force is used reduce the relative motion of ice elements passing by one another with overlapping boundaries.
%The perpendicular damping force is used to damping coefficient in the the perpendicular direction is set to $c_{r_{\perp}}=\frac{1}{4}c_{r_{||}}$. 
%The frictional damping force perpendicular to $\vec{r}_{ij}$ helps to reduce the motion of particles when many particles are packed tightly together (as in the uncoupled experiments below), and also acts to reduce the vibrations within larger bonded structures. 
The damping forces are implemented using an implicit time stepping scheme, to avoid stability issues for very small elements (details found in Appendix B).
